Adaptive evolution with aneuploidy and mutation

### 2022-01-26 ms/ms.tex
added MP text from a analysis branch and revised it, added todos

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{subcaption}
\captionsetup[subfigure]{labelformat=empty}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{longtable}
\usepackage{gensymb,siunitx}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Evolutionary Models.}
We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\autoref{fig:models}), as well as for a fluctuating population size.

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Single-locus model.}
This model assumes a constant effective population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.

Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \autoref{table:single-locus}.

@@

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.

@@

The change in frequencies due to selection is exactly the same as in the single-locus model (\autoref{eq:selection-single}), only applied using the fitness values in \autoref{table:multi-locus}. The change due to random genetic drift is also the same as in \autoref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.

@@

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
\autoref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.

@@

The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below). The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. 
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
$K_t(\theta' \mid \theta)$ is a multivariate normal distribution, fitted at iteration $t$ to the particles from the previous iteration, \{$\theta_{i,t-1}\}_{i=1}{n_{t-1}}$, and their weights, $\{W(\theta_{i,t-1})\}_{i=1}^{n_{t-1}}$.

Acceptance is determined according to the approximate likelihood~(\autoref{eq:heatstress-likelihood}), which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\autoref{fig:convergence}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.

@@

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\autoref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\autoref{fig:growth-curves}-A). In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
%In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. Such high fitness values are unreasonable. Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27. 
% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
% TODO what are the results ?
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\autoref{fig:growth-curves}-B) to estimate $1-c$ . This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO it make sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\autoref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). The estimated parameter values without aneuploidy are: 
%% generated with MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{subcaption}
\captionsetup[subfigure]{labelformat=empty}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{longtable}
\usepackage{gensymb,siunitx}
\usepackage{cleveref}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Evolutionary Models.}
We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\cref{fig:models}), as well as for a fluctuating population size.

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Single-locus model.}
This model assumes a constant effective population size $N$ and follows four genotypes~(\cref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.

Transitions between the genotypes occur as follows~(\cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \cref{table:single-locus}.

@@

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

The fitness of the different genotypes is the same as in the single-locus model (\cref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \cref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.

@@

The change in frequencies due to selection is exactly the same as in the single-locus model (\cref{eq:selection-single}), only applied using the fitness values in \cref{table:multi-locus}. The change due to random genetic drift is also the same as in \cref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.

@@

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
\cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.

@@

The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below). The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. 
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
$K_t(\theta' \mid \theta)$ is a multivariate normal distribution, fitted at iteration $t$ to the particles from the previous iteration, \{$\theta_{i,t-1}\}_{i=1}{n_{t-1}}$, and their weights, $\{W(\theta_{i,t-1})\}_{i=1}^{n_{t-1}}$.

Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\cref{fig:convergence}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\cref{fig:seeds}). %TODO for multi-locus model, write about them here also.

@@

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-A). In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
%In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. Such high fitness values are unreasonable. Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27. 
% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
% TODO what are the results ?
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ . This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO it make sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). The estimated parameter values without aneuploidy are: 
%% generated with MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?
